---
title: "Exploratory Analysis: the Opioid Crisis and its relationship with the Pharmaceutical Industry"
author: "Team 2 INFO 478"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("pharma.R")
source("income.R")
source("overdoses.R")
```

>  For our population health informatics project, our group will be focusing on the second wave of the United States opioid epidemic (starting around 2010). The first wave of the opiod crisis began in the mid-1990's when the pharmacutical drug OxyContin was developed and mass produced by Purdue Pharma, and the second wave began roughly in 2010 and lasted for a few years. The mortality of the second wave is intensified due to the growing heroin market which attracted populations who were already addicted to perscription opioids in addition to those who developed a prescription drug addiction. Most recently, starting in 2018, the United States has begun to see a third wave with deaths surging from synthetic opioids like fentanyl entering the illegal drug market. With our focus being the second wave, our variable of interest is the financial success of major US pharmacutica companies who, under minimal regulations, contributed to the severity of the epidemic. The health metric that we will use for our visualizations will be mortality (sum of all opioid related deaths in a calendar year in the US). The opioid epidemic is not only a serious public heath crisis, but is also extremely damaging to the US economy, with $78.5 billion in losses each year due to lost productivity, addiction treatment, and criminal justice involvement.



## Visualization of Opioid Related Deaths and State Household Income 

``` {r income, warning=FALSE, echo=FALSE}
income_vs_deaths_2017_plot
```

> From the scatterplot given above, the trend between total number of opioid related-deaths and average household income can be explored across different states in 2017. The states with the highest average household incomes are Maryland, Washington, and New Hampshire, and the states with the highest number of opioid deaths are Pennsylvania, Ohio, and Florida. Those with the lowest average household incomes are Mississippi, Louisiana, and West Virginia, and those with the lowest number of opioid deaths are North Dakota, Wyoming, and South Dakota. From these data points alone, there seems to be no direct correlation between average household income and opiod deaths. However, there could be other overarching correlations in the data that skew the trend. For instance, the scatter plot charts the relationship using the total number of opioid-related deaths, not the standardized rate of deaths per population size. By plotting deaths using the total number metric instead of a rate, the visualization allows for the possibility of other trends that cannot be attributed to household income. Two states that we will examine more closely are Maryland and Massachusetts, since these states have higher incomes and a high amount of overdose deaths. 


## Map of the Rate of Opiod Overdoses by State

``` {r overdoses, warning=FALSE, echo=FALSE}
od_rate_2019_map
```

> This heat map visually compares the rate of opiod overdose deaths across different states in 2019. The metric used to measure rate was the *age-adjusted rate per population of 100,000*, which standardizes the rate of death between states with different age distributions and overall populations. The map indicates that the state with the highest rate of death caused by opiod overdose was **West Virginia**, with about 52 deaths per 100,000, while the state with the lowest rate of death by overdose was **Nebraska**, with about 9 deaths per 100,000. The distribution of color in the heat map also signifies that overdose deaths were most prevalent in the Midwest, with high rates in states such as West Virginia, Ohio, Maryland, Pennsylvania, and Kentucky. Regions of the country with lower rates of opiod overdoses included the Central U.S. (Nebraska, South Dakota, and Iowa), and the South (Texas, Alabama, and Georgia).

## Top 10 States with most OD Deaths (1999 - 2014)
``` {r top10, warning=FALSE, echo=FALSE}
bar_chart_deaths
```

>In the box plot above, we can see the top ten U.S. states with the most opioid overdose deaths from 1999-2014. These deaths are not age-adjusted and are the sum of all deaths that occurred in that state with the cause of death being an overdose relating to either illegal opioids (like heroin and fentanyl) or prescription drug abuse. Of these states, the ones with the highest amount of deaths are **California**, 
**Texas**, and **Florida**. These states all have high populations; for reference the most populated states in the U.S. are California, Texas, Florida, New York, and Pennsylvania. Other states that are less populated but appear on the graph are **Washington**, **Maryland**, **Massachusetts**, and **North Carolina**. The top three states' deaths can be attributed to their population density, but these other states making the top ten list may have other factors attributing to their high amount of deaths. From research on the opioid crisis through all its waves, we know that North Carolina, Ohio, and Maryland were all severely impacted by the opioid crisis and that these states have higher use of both illegally manufactured opioids and prescription drug abuse. From this top ten list we can infer that the opioid crisis has made a devastating effect on many U.S. states, and that states with higher populations are more likely to have citizens who misuse this class of drug. Interventions designed to prevent these overdose deaths are crucial in all U.S. states but would have the greatest impact on these states. 



## Prescriptions Dispensed vs Deaths (1999 - 2014)
``` {r prescriptions, warning=FALSE, echo=FALSE}
scatter_pres_deaths
```

> In the scatterplot of prescriptions dispensed and deaths, we can observe a strong positive correlation between prescriptions dispensed for pain relief drugs (opioids) and overdose deaths. Each point along the graph represents a different year, and the most devastating years are the ones where prescriptions dispensed are over 1000 million. This strong correlation demonstrates that when physicians over prescribe pain relief medication without proper warnings on its risks and addictive qualities, overdose deaths increase. The years with the highest amount of opioids prescribed and overdose deaths are years where prescription opioids, like hydrocodone and oxymorphone, were used to treat pain management conditions and before U.S. regulatory agencies addressed the risks of these drugs. This scatterplot further suggests that if prescriptions dispensed for these drugs were to decrease, that we would likely see a decrease in overdose deaths. Public health experts can use this information to guide their practices surrounding prescribing these drugs and limiting their misuse among patients. 
