---
title: "Exploratory Analysis: the Opioid Crisis and its reltionship with the Pharmacutical Industry"
author: "Team 2 INFO 478"
date: "2/22/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("pharma.R")
source("income.R")
```

> For our population health informatics project, our group will be focusing on the second wave of the United States opioid epidemic (starting around 2010). The first wave of the opiod crisis began in the mid-1990's when the pharmacutical drug OxyContin was developed and mass produced by Purdue Pharma, and the second wave began roughly in 2010 and lasted for a few years. The mortality of the second wave is intensified due to the growing heroin market which attracted populations who were already addicted to perscription opioids in addition to those who developed a prescription drug addiction. Most recently, starting in 2018, the United States has begun to see a third wave with deaths surging from synthetic opioids like fentanyl entering the illegal drug market. With our focus being the second wave, our variable of interest is the financial success of major US pharmacutica companies who, under minimal regulations, contributed to the severity of the epidemic. The health metric that we will use for our visualizations will be mortality (sum of all opioid related deaths in a calendar year in the US). The opioid epidemic is not only a serious public heath crisis, but is also extremely damaging to the US economy, with $78.5 billion in losses each year due to lost productivity, addiction treatment, and criminal justice involvement. 

## Visualization of Opiod Related Deaths and the Pharmaceutical Industry's Revenues

``` {r pharma, warning=FALSE, echo=FALSE}
comp_scatter
```

> In the scatterplot above we examine the relationship between total opiod related deaths (nationwide mortlity) and the total revenue of the 12 largest pharmacutical companies from the years 2009-2012. In the scatterplot we see a strong positive correlation between deaths and revenues. This effect may have other confounding factors, such as other prescription drugs like insulin increasing in cost, or economic trends, but we do see that as opioid and heroin deaths increase, the pharmacutical industry is not negatively impacted by these deaths. This shows that these large multinational corporations are not seeing consquences of the misuse and over prescription of pain relievers financially. Over this time interval, we see that the most deaths occured in 2011, with `r deaths_2011` deaths total due to the opioid crisis. The highest revenues for the pharmacutical industry occured in 2012, with revenue being `r revenue_2012` million dollars. 

## Visualization of Opiod Related Deaths and State Household Income 

``` {r income, warning=FALSE, echo=FALSE}
income_vs_deaths_2017_plot
```

